---
title: "in romana"
output:
  html_document: default
  word_document: default
date: "2024-11-26"
---

---
title: "workflow"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
date: "2024-08-09"
---

```{r setup, include=FALSE}
library(readxl)
library(FactoMineR)
library(factoextra)
library(NbClust)
library(rgl)
library(pixiedust)
library(multcomp)
library(car)
library(nlme)
library(psych)
library(corrplot)
library(htmlwidgets)
library(MASS)
library(effectsize)
library(cluster)
library(fpc)
library(tidyverse)
library(psych)
library(janitor)
library(patchwork)
library(viridis)
library(stargazer)

set.seed(2024)
theme_set(theme_light())

Razboieni_pXRF <- read_excel("Razboieni_pXRF.xlsx", 
    sheet = "Selectie RSD<10")
```

#Data Selection
##Removing variables


Remove from analysis the elements that are troublesome for quantitative analysis (Phosphorus per Hunt & Speakman 2015 JAS 53)

```{r}
df <- as.data.frame(Razboieni_pXRF)
df[,1:2] <- lapply(df[,1:2], as.factor)
df <- df |> 
  unite("Proba_Tip", Tip:SAMPLE, sep = "_", remove = F) |> 
  column_to_rownames("Proba_Tip") |> 
  unite("Proba_Tip", Tip:SAMPLE, sep = "_", remove = F)

```

#EDA

```{r}
library(GGally)
ggpairs(df, columns = 4:14) +
  ggtitle("Grafic 1. Analiza Exploratorie")
  

```


```{r}
#df <- df[, -c(4,6)]
df_check_point <- df
```


#Data Selection
##Removing individuals
###Multivariate Outlier Analysis and Elimination using Local Outlier Factor


```{r}
library(dbscan)
lof_scores <- dbscan::lof(scale(df[,4:14], center = T, scale = T), minPts = 4)
summary(lof_scores)
df$lof_scores <- lof_scores
ggplot(df, aes(x = 1:nrow(df), y = lof_scores)) +
  geom_point() + 
  geom_text(aes(label = Proba_Tip)) +
  theme_light() +
  xlab(label = "") +
  ylab(label = "Scorul Densității Locale") +
  ggtitle("Grafic 2. Identificarea valorilor extreme")
```
From the graph it is clear that samples 01, 16, 17, are outliers and will be removed from the analysis as they may impact the clustering results

```{r}
threshold <- 1.5
df$outlier <- ifelse(lof_scores > threshold, "Valori Extreme", "Inlier")
df <- filter(df, !df$outlier == "Valori Extreme")
ggpairs(df, columns = 4:14) +
  ggtitle("Grafic 3. Analiza Exploratorie după eliminarea valorilor extreme")

```

#Clustering

##Calculate clustering tendency

```{r}
gradient_col <- list(low = "black", mid = "white",high = "red")


df_centr_clust <- as.data.frame(scale(df[,4:14],
                                               center = TRUE,
                                               scale = TRUE)
                                         )
library(hopkins)
hopkins::hopkins(df_centr_clust, m = nrow(df) - 2)

fviz_dist(dist(df_centr_clust), show_labels = T) +
  labs(title = "Grafic 4. Tendințe de grupare",
       subtitle = "Indice Hopkins: 0.959638") 
  
```

##Agglomerative Hierachical Clustering

```{r}
clust_attr <- eclust(df_centr_clust,                       
                                      FUNcluster = "agnes",             
                                      hc_metric = "minkowski",         
                                      hc_method = "ward.D2")           
clusts_indet <- fviz_dend(clust_attr) +
  ggtitle("Grafic 5. Dendrograma clusterelor")
clusts_indet

library(clValid)
distance_matrix <- distance(df_centr_clust)
distance_matrix <- as.matrix(distance_matrix)
internal_validation <- clValid(distance_matrix, 2:8, 
                               clMethods = c("hierarchical"),
                               validation = c("internal", "stability"))
summary(internal_validation)
```

The dendrogram indicates that the greatest distances between grpups exist if these are grouped in three clusters

```{r}
clust_2 <- eclust(df_centr_clust,                       
                                      FUNcluster = "agnes",             
                                      k = 5,      
                                      hc_metric = "minkowski",         
                                      hc_method = "ward.D2")           

silh_2 <- fviz_silhouette(clust_2) #+
  labs(title = "Gafic 6. Indicele siluetei pentru partiționări în clustere",
       subtitle = "Partiționare in trei clustere. Indicele mediu al siluetei: 0,25")
clusts_tests_2 <- fviz_dend(clust_2, rect = T, rect_fill = T)
clusts_tests_2


clust_7 <- eclust(df_centr_clust,                       
                                      FUNcluster = "agnes",             
                                      k = 7,      
                                      hc_metric = "minkowski",         
                                      hc_method = "ward.D2")           



silh_7 <- fviz_silhouette(clust_7) #+
  labs(title = "Gafic 6. Indicele siluetei pentru partiționări în clustere",
       subtitle = "Partiționare in șapte clustere. Indicele mediu al siluetei: 0,25")
clusts_tests_7 <- fviz_dend(clust_7, rect = T, rect_fill = T)
clusts_tests_7

clust_8 <- eclust(df_centr_clust,                       
                                      FUNcluster = "agnes",             
                                      k = 8,      
                                      hc_metric = "minkowski",         
                                      hc_method = "ward.D2")           


silh_2 / silh_7

```

The silhouette index is low, therefore further verification of clustering accuracy is necessary

Completion of data with the cluster assignments

```{r}
clusteri_2 <- clust_2$cluster
df$clusters_2 <- as.factor(clusteri_2)
clusteri_7 <- clust_7$cluster
df$clusters_7 <- as.factor(clusteri_7)
```


#Permutation test for linear model

```{r}
library(lmPerm)

model_2_clusters <- aovp(Al2O4_avg + SiO3_avg + K2O_avg + Ca_avg + Ti_avg + Mn_avg + Fe_avg + Zn_avg + Rb_avg + Sr_avg + Zr_avg ~ clusteri_2,
              seqs = F,
              maxIter = 1e6,
              data = df, 
              perm = "Exact")
summary.aovp(model_2_clusters)


model_7_clusters <- aovp(Al2O4_avg + SiO3_avg + K2O_avg + Ca_avg + Ti_avg + Mn_avg + Fe_avg + Zn_avg + Rb_avg + Sr_avg + Zr_avg ~ clusteri_7,
              seqs = F,
              maxIter = 1e6,
              data = df, 
              perm = "Exact")
summary.aovp(model_7_clusters)

model_2_elements <- aovp(cbind(Al2O4_avg, SiO3_avg, K2O_avg, Ca_avg, Ti_avg, Mn_avg, Fe_avg, Zn_avg, Rb_avg, Sr_avg, Zr_avg) ~ clusters_2,
             seqs = F,
             maxIter = 1e6,
             data = df, 
             perm = "Exact")
summary.aovp(model_2_elements)


model_7_elements <- aovp(cbind(Al2O4_avg, SiO3_avg, K2O_avg, Ca_avg, Ti_avg, Mn_avg, Fe_avg, Zn_avg, Rb_avg, Sr_avg, Zr_avg) ~ clusters_7,
             seqs = F,
             maxIter = 1e6,
             data = df, 
             perm = "Exact")
summary.aovp(model_7_elements)


```



#PCA with clusters overlay

```{r}
pca_res <- PCA(df[,4:14], graph = FALSE, scale.unit = T)
summary(pca_res)

plot_pca <- fviz_pca_biplot(pca_res, addEllipses = T, ellipse.level = 0.90,
                                     col.ind = df$clusters_2,
                                     geom.ind = "point",
                                     col.var = "black",                       
                                     repel = TRUE,                            
                                     title = "Grafic 8. Biplot CP 1 & CP 2 cu indicarea clusterelor",
                                     legend.title = "Clustere") + 
  geom_text(label = df$Proba_Tip, hjust = 0.5, nudge_y = -0.09, check_overlap = TRUE) +
   scale_colour_viridis_d()

plot_pca
```

######PCA - visualisations - eigen values
```{r}
eigen_values <- pca_res$eig |> 
  as_tibble() |> 
  rownames_to_column() |> 
    janitor::clean_names() |> 
  arrange(desc(percentage_of_variance)) 


graph_eigen <- eigen_values |> 
    top_n(8, percentage_of_variance) |>
  mutate(percentage_of_variance = round(percentage_of_variance, 2)) |> 
  ggplot(aes(rowname, percentage_of_variance)) +
  geom_col() +
  geom_text(aes(label = percentage_of_variance, vjust = 1.5,
              )) +
  xlab(label = "Componente Principale") +
  ylab(label = "Procentajul variației Explicate") +
  labs(title = "Grafic 7. Puterea de explicație a Componentelor Principale Identificate")
graph_eigen
```

######PCA - visualisations - variables
```{r}
contrib_var_selection <- pca_res$var$contrib |> 
  as.data.frame() |> 
  rownames_to_column() |> 
  mutate_if(is.character, as.factor) 

graph_dim1_contrib <- 
  contrib_var_selection |> 
  mutate(rowname = fct_reorder(rowname, Dim.1)) |> 
  ggplot(aes(Dim.1, rowname)) +
  geom_col() +
  xlab(label = "Contribuții la CP 1") +
  ylab(label = "")


graph_dim2_contrib <- 
  contrib_var_selection |> 
  mutate(rowname = fct_reorder(rowname, Dim.2)) |> 
  ggplot(aes(Dim.2, rowname)) +
  geom_col() +
  xlab(label = "Contribuții la CP 2") +
  ylab(label = "")


graph_dim3_contrib <- 
  contrib_var_selection |> 
  mutate(rowname = fct_reorder(rowname, Dim.3)) |> 
  ggplot(aes(Dim.3, rowname)) +
  geom_col() +
  xlab(label = "Contribuții la CP 3") +
  ylab(label = "")

cors_var <- pca_res$var$cor |> 
  as.data.frame() |> 
  rownames_to_column() |> 
  mutate_if(is.character, as.factor) 

graph_dim1_cors <- 
  cors_var |> 
  mutate(rowname = fct_reorder(rowname, Dim.1, .desc = T)) |> 
  ggplot(aes(Dim.1, rowname)) +
  geom_col() +
  theme(axis.text = element_text(size = 9, face = "bold"),
        axis.title = element_text(face = "bold")
        ) + 
  xlab(label = "Corelația Elementelor cu CP 1") +
  ylab(label = "") +
  inset_element(graph_dim1_contrib,
                left = 0.8,
                bottom = 0.5,
                right = 1,
                top = 1)

graph_dim2_cors <- 
  cors_var |> 
  mutate(rowname = fct_reorder(rowname, Dim.2, .desc = T)) |> 
  ggplot(aes(Dim.2, rowname)) +
  geom_col() +
  theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 9, face = "bold")
        ) + 
  xlab(label = "Corelația Elementelor cu CP 2") +
  ylab(label = "") +
  inset_element(graph_dim2_contrib,
                left = 0.8,
                bottom = 0.5,
                right = 1,
                top = 1)


graph_dim3_cors <- 
  cors_var |> 
  mutate(rowname = fct_reorder(rowname, Dim.3, .desc = T)) |> 
  ggplot(aes(Dim.3, rowname)) +
  geom_col() +
  theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(size = 9, face = "bold")
        ) + 
  xlab(label = "Corelația Elementelor cu CP 3") +
  ylab(label = "") +
  inset_element(graph_dim3_contrib,
                left = 0.8,
                bottom = 0.5,
                right = 1,
                top = 1)
```

######PCA - visualisations - cases
```{r}
coord_ind <- 
  pca_res$ind$coord |> 
  as.data.frame() |> 
  mutate_if(is.character, as.factor) |> 
  rownames_to_column() |> 
  as_tibble() |> 
  separate_wider_delim(rowname, delim = "_", 
                       names = c("Tip", "Proba", "Numar"),
                       cols_remove = F)

graph_ind_1 <- coord_ind |> 
  mutate(rowname = fct_reorder(rowname, Dim.1, .desc = T)) |> 
  ggplot(aes(Dim.1, rowname)) +
  geom_col(aes(colour = df$clusters, fill = df$clusters_2)) +
  theme(legend.position = "inside",
    legend.position.inside = c(1,1),
    legend.justification.inside = c(1,1),
    axis.text = element_text(size = 9, face = "bold"),
    axis.title = element_text(face = "bold")
    ) + 
  xlab(label =  "Corelația Probelor cu CP 1") + 
  ylab(label =  "Probe") +
  scale_fill_viridis(discrete = T) +
  scale_color_viridis_d() +
  guides(fill = guide_legend(title = "Clustere")) +
  guides(color = F)

graph_ind_2 <- coord_ind |> 
  mutate(rowname = fct_reorder(rowname, Dim.2, .desc = T)) |> 
  ggplot(aes(Dim.2, rowname)) +
  geom_col(aes(colour = df$clusters_2, fill = df$clusters_2)) +
  theme(legend.position = "inside",
    legend.position.inside = c(1,1),
    legend.justification.inside = c(1,1),
    axis.text = element_text(size = 9, face = "bold"),
    axis.title = element_text(face = "bold")
    ) + 
  xlab(label =  "Corelația Probelor cu CP 2") + 
  ylab(label =  "Probe") +
  scale_fill_viridis(discrete = T) +
  scale_color_viridis_d() +
  guides(fill = guide_legend(title = "Clustere")) +
  guides(color = F)

graph_ind_3 <- coord_ind |> 
  mutate(rowname = fct_reorder(rowname, Dim.3, .desc = T)) |> 
  ggplot(aes(Dim.3, rowname)) +
  geom_col(aes(colour = df$clusters_2, fill = df$clusters_2)) +
  theme(legend.position = "inside",
    legend.position.inside = c(1,1),
    legend.justification.inside = c(1,1),
    axis.text = element_text(size = 9, face = "bold"),
    axis.title = element_text(face = "bold")
    ) + 
  xlab(label =  "Corelația Probelor cu CP 3") + 
  ylab(label =  "Probe") +
  scale_fill_viridis(discrete = T) +
  scale_color_viridis_d() +
  guides(fill = guide_legend(title = "Clustere")) +
  guides(color = F)
```

PCA - bring it home
```{r}
graph_CP1 <- graph_dim1_cors / graph_ind_1 +
  plot_annotation(title = "Grafic 9. Definirea clusterelor în funcție de CP 1")
graph_CP2 <- graph_dim2_cors / graph_ind_2 +
  plot_annotation(title = "Grafic 10. Definirea clusterelor în funcție de CP 2")
graph_CP3 <- graph_dim3_cors / graph_ind_3 +
  plot_annotation(title = "Grafic 11. Definirea clusterelor în funcție de CP 3")
graph_CP1
graph_CP2
graph_CP3
```

The first component defines the difference between fine and coarse pottery and between clusters 4&2 and 1&3

The second component defines the difference between the clusters 2&4, and is probably expressing differences in the raw materials.




# Biplot of Ti vs Zr
The distribution of the elemental ratio of Ti/Zr (titanium/zirconium) can be used to indicate heterogeneity of parent materials because they weather slowly and can retain the characteristics of parent materials (Encyclopedia of Soils in the Environment). 

```{r}
df_ti_zr <- full_join(df, df_check_point) 
df_ti_zr$clusters_2 <- coalesce(df_ti_zr$clusters_2, "Valori Extreme")
df_ti_zr$clusters <- as.factor(df_ti_zr$clusters_2)

ggplot(df_ti_zr, aes(x = Ti_avg, y = Zr_avg, color = clusters_2)) +
  geom_point(show.legend = T) +
  geom_text(label = df_ti_zr$Proba_Tip, hjust = 0.0005, nudge_y = -0.0005, check_overlap = TRUE, show.legend = F) +
    labs(colour = "Legenda",
         title = "Grafic 12. Exprimare grafcă a raportului Zr/Ti") #+
  scale_color_viridis_d()
  
df_ti_zr_7 <- full_join(df, df_check_point) 
df_ti_zr_7$clusters_7 <- coalesce(df_ti_zr$clusters_7, "Valori Extreme")
df_ti_zr_7$clusters <- as.factor(df_ti_zr_7$clusters_7)

ggplot(df_ti_zr_7, aes(x = Ti_avg, y = Zr_avg, color = clusters_7)) +
  geom_point(show.legend = T) +
  geom_text(label = df_ti_zr_7$Proba_Tip, hjust = 0.0005, nudge_y = -0.0005, check_overlap = TRUE, show.legend = F) +
    labs(colour = "Legenda",
         title = "Grafic 12. Exprimare grafcă a raportului Zr/Ti") +
  scale_color_viridis_d()

```
